A generalization of the Bradley-Terry model for draws in chess

### 2019-05-14 inst/collusion_chess_draws.Rnw
work

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess match is
regarded as a three-way competition between the two players and an
entity that wins if the match is drawn.  Bradley-Terry can be further
modified to account for the White player advantage in a number of
ways.  This methodology affords insight into players' strengths,
response to playing Black or White, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.}

@@

\cite{moul2009}

@@

moul: White's relative advantage over Black will be dependent upon the
comparative mastery of the two players.

The most obvious issue regarding the threshold against which we compare
this relative advantage of White is the inherent advantage of White's
position in moving first. We correspondingly do not expect these base
thresholds to be symmetric and impose no such restriction.

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be to two-way
competitive situations such as tennis in which one or other player is
idenfied as the winner, but the system has been applied to many
non-competitive situations such mating preference~\cite{hankin_aylmer}
and consumer behaviour.  The Bradley-Terry model~\citep{bradley1952}
considers $n$ players and assigns non-negative numbers
$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
has been modified and extended many times, for example to account for
multi-way choices such as icon selection~\cite{hankinhyperdirichlet}.

@@

Chess games may be won by either player, or may be drawn. The
implication of a draw is that neither player is better than the other.
Before 1867, tournament games that were drawn were replayed and in
this context we note that this procedure reduces a chess match to a
Bernoulli trial, with probability one.  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn match half a point, rendering chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are generally poorer in spectacular or striking elements such
as sharp tactical combinations or sacrifice.  Excessive drawing
dissuades sponsors, leaves spectators feeling cheated, and leaves
commentators with material to discuss with the public.
\cite{moul2009} concurs, conceptualising a player as being able to
adopt a strategy on a spectrum ranging from conservative to
aggressive, thus being able to control the probability of a draw.  He
suggests that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance.

such as chess, the Bradley-Terry model can
be augmented to chess-specific considerations in a number of ways.
Here I present a new generalization of Bradley-Terry in which a chess
match is regarded as a three-way competition between the two players
and an entity that wins if the match is drawn.  Bradley-Terry can be
further modified to account for the White player advantage in a number
of ways.  This methodology affords insight into players' strengths,
response to playing Black or White, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.

@@

where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
the number of times $i$ loses to $j$.  Incorporation of draws is
straightforward: we consider a chess match to be a three-way trial
between the two players and an entity that wins if the match is drawn.
We thus seek a likelihood function for non-negative
$p_1,\ldots,p_n,p_D$ with $p_D+\sum p_i=1$ and this would be

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by playing white.  We
thus consider a chess match between two players to be a three-way
competition between the two ``real'' players and the draw monster, but
this time we interpret the white player as the combined strength of
the player himself and another entity, the White monster, whose
strength $p_W$ is added to the white player.

@@

\section{Chess collusion: Cura\c{c}ao 1962}
\begin{table}
\centering
\begin{tabular}{ll|cccccccc}
 &
&\rotatebox{-90}{Petrosian} 
&\rotatebox{-90}{Keres} 
&\rotatebox{-90}{Geller}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Korchnoi}
&\rotatebox{-90}{Benko}
&\rotatebox{-90}{Tal}
&\rotatebox{-90}{Filip}\\
Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &A \\
\end{tabular}
\caption{Cura\c{c}ao 1962 candidates matches, world championship 1963}
\end{table}

>>>>

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess match is
regarded as a three-way competition between the two players and an
entity that wins if the match is drawn.  Bradley-Terry can be further
modified to account for the White player advantage in a number of
ways.  This methodology affords insight into players' strengths,
response to playing Black or White, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.

The likelihood functions arising are easily optimised numerically.  I
analyse a number of datasets of chess results, including the infamous
World Chess Championships 1963 in which the Soviets were accused of
collusion.

}

@@

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be to two-way
competitive situations such as tennis in which one or other player is
idenfied as the winner, but the system has been applied to many
non-competitive situations such mating preference~\citep{west2008} and
consumer behaviour.  The Bradley-Terry model~\citep{bradley1952}
considers $n$ players and assigns non-negative numbers
$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
has been modified and extended many times, for example to account for
multi-way choices such as mating preferences~\citep{hankin2010}.

@@

Chess games may be won by either player, or may be drawn. The
implication of a draw is that neither player is better than the other.
Before 1867, tournament games that were drawn were replayed and in
this context we note that this procedure reduces a chess match to a
Bernoulli trial, with probability one.  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn match half a point, rendering chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citeauthor{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
34\% of games are drawn ($\chi^2$, $p<3\times 10^{-6}$).

\cite{moul2009} conceptualise a player as being able to adopt a
strategy on a spectrum ranging from conservative to aggressive, thus
being able to exert some control over the probability of a draw.  They
suggest that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance. \cite{langer2013} {\bf
  cite something better here} point out that rational players in an
ongoing competitive situation modify their strategies depending on
whether they are leading or lagging: the lagging player will adopt
higher risk strategy (`nothing to lose'), while the leading player is
incentivised to protect the lead by lowering the risk of losing.  In a
chess context, this means the leading player will aim for a draw.

Such arguments suggest that considering a drawn chess match as a
qualitatively different outcome from a win for either player: the play
itself tends to be different, the drawing players have a different
mindset, and drawn games have less value in the wider context of the
game.

\section{Bradley-Terry}

@@

where $w_{ij}$ is the number of times $i$ beats $j$ and $l_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess match is regarded as
a three-way competition between the two players and an entity that
wins if the match is drawn.  The system is further modified to account
for first-move advantage by positing an entity whose strength is added
to that of the White player.  This methodology affords insight into
players' strengths, response to playing Black or White, and
risk-aversion as manifested by probability of drawing.

If we consider a chess match to be a three-way trial, we define a
likelihood function for non-negative $p_1,\ldots,p_n,p_D$ with
$p_D+\sum p_i=1$ and this would be

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength is added to
the white player.  The strength of the white player is thus the sum of
the player's strength and another entity, the White monster.
Recalling that a game might be a draw, we obtain the probabilities as
given in Table~\ref{probabilities}.  The corresponding likelihood
function is given below in equation \ref{likelihood_WD}:

\newcommand{\oo}{\vphantom{W^{W^W}}}

\begin{table}
\centering
\begin{tabular}{l|cc}\\
  &$i$ plays white & $j$ plays white\\ \hline
$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+p_W\oo}{p_i+p_j+p_W+D}$ & $\frac{p_i}{p_i+p_j+p_W+D}$  \\
$\Prob{(\mbox{draw})}$      & $\frac{D\oo}{p_i+p_j+p_W+D}$       & $\frac{D}{p_i+p_j+p_W+D}$    \\
$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+p_W+D}$     & $\frac{p_j+p_W}{p_i+p_j+p_W+D}$\\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two players under a generalized Bradley-Terry model\label{probabilities}}
\end{table}

@@

\section{Chess collusion: Cura\c{c}ao 1962}

\begin{table}
\centering
\begin{tabular}{ll|cccccccc}
 &
&\rotatebox{-90}{Petrosian} 
&\rotatebox{-90}{Keres} 
&\rotatebox{-90}{Geller}
&\rotatebox{-90}{Fischer} 
&\rotatebox{-90}{Korchnoi}
&\rotatebox{-90}{Benko}
&\rotatebox{-90}{Tal}
&\rotatebox{-90}{Filip}\\
Petrosian &USSR    &             &DDDD    &DDDD   &DWDD   &DDWW   &DDWD   &WWD    &DWWD\\
Keres     &USSR    &             &NA      &DDDD   &LDWD   &DDWD   &WWWL   &WDW    &DWWD\\
Geller    &USSR    &             &NA      &NA     &WWDL   &DDWD   &DDDW   &DWW    &DWWD\\
Fischer   &USA     &             &NA      &NA     &NA     &LWLD   &LWDW   &DWD    &WDWD\\
Korchnoi  &USSR    &             &NA      &NA     &NA     &NA     &DDDL   &WLD    &WWWW\\
Benko     &USA     &             &NA      &NA     &NA     &NA     &NA     &WLD    &LWWD\\
Tal       &USSR    &             &NA      &NA     &NA     &NA     &NA     &NA     &WLD \\
Filip     &TCH     &             &NA      &NA     &NA     &NA     &NA     &NA     &A \\
\end{tabular}
\caption{Cura\c{c}ao 1962 candidates matches, world championship 1963}
\end{table}
